```{r}
# 0. Setup ----
required_pkgs <- c("readr","dplyr","ggplot2","GGally","corrplot",
                   "car","MASS","gridExtra","scales")
missing_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[,"Package"])]
if (length(missing_pkgs)) {
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
}
library(readr); library(dplyr); library(ggplot2); library(GGally); library(corrplot)
library(car); library(MASS); library(gridExtra); library(scales)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
data_path <- "ENB.txt"
```

```{r}
# 1. Read data ----
raw <- read_table2(data_path, col_names = FALSE, skip = 1, progress = FALSE)
is_num_col <- sapply(raw, is.numeric)
df_num <- raw[, is_num_col]
if (ncol(df_num) > 6) {
  df_num <- df_num[ , (ncol(df_num)-5):ncol(df_num)]
}
colnames(df_num) <- c("X1","X2","X3","X4","X5","Y")
df <- as.data.frame(df_num)
dim(df)
head(df)
```

There are 6 variables (5 predictors X1 to X5 and 1 response Y) in the dataset. Total observations are 19735.

```{r}
# 2. Basic summary / missing values / types ----
summary_stats <- summary(df)
summary_stats
sapply(df, function(x) sum(is.na(x)))
```

```{r}
# 3. Univariate plots (distribution) ----
plot_list <- list()
for (v in names(df)) {
  p_hist <- ggplot(df, aes_string(x = v)) +
    geom_histogram(aes(y = ..density..), bins = 30, fill = "steelblue", color = "white") +
    geom_density(alpha = 0.3, color = "red") +
    labs(title = paste("Histogram & density -", v)) +
    theme_minimal()
  p_box <- ggplot(df, aes_string(y = v)) +
    geom_boxplot(fill = "orange", alpha = 0.7) +
    labs(title = paste("Boxplot -", v)) +
    theme_minimal()
  plot_list[[paste0(v,"_hist")]] <- p_hist
  plot_list[[paste0(v,"_box")]] <- p_box
  print(p_hist)
  print(p_box)
}
```

Variables X1, X2,X5 and Y show potential outliers in boxplots. Right skewness is observed in X1,Y distributions. Left skewness is observed in X3 and X5 and

```{r}
# 4. Pairwise relationships ----
GGally::ggpairs(df, columns = 1:6, progress = FALSE)
```
#scatter plots 
```{r}
# Choose the source matrix
src <- df  # or use the.data

# If there is a leading index column, drop it
if (ncol(src) == 7) src <- src[, 2:7, drop = FALSE]

# Ensure column names exist; assign if needed
if (is.null(colnames(src)) || !all(colnames(src) %in% c("X1","X2","X3","X4","X5","Y"))) {
  colnames(src) <- c("X1","X2","X3","X4","X5","Y")
}

# Set a 2x3 layout for five plots
op <- par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))

plot(src[,"X1"], src[,"Y"], xlab = "X1: Living room temp (°C)", ylab = "Y: Appliances (Wh)",
     pch = 19, col = rgb(0,0,1,0.4), main = "X1 vs Y")
abline(lm(Y ~ X1, data = as.data.frame(src)), col = "red", lwd = 2)

plot(src[,"X2"], src[,"Y"], xlab = "X2: Living room humidity (%)", ylab = "Y: Appliances (Wh)",
     pch = 19, col = rgb(0,0,1,0.4), main = "X2 vs Y")
abline(lm(Y ~ X2, data = as.data.frame(src)), col = "red", lwd = 2)

plot(src[,"X3"], src[,"Y"], xlab = "X3: Office temp (°C)", ylab = "Y: Appliances (Wh)",
     pch = 19, col = rgb(0,0,1,0.4), main = "X3 vs Y")
abline(lm(Y ~ X3, data = as.data.frame(src)), col = "red", lwd = 2)

plot(src[,"X4"], src[,"Y"], xlab = "X4: Office humidity (%)", ylab = "Y: Appliances (Wh)",
     pch = 19, col = rgb(0,0,1,0.4), main = "X4 vs Y")
abline(lm(Y ~ X4, data = as.data.frame(src)), col = "red", lwd = 2)

plot(src[,"X5"], src[,"Y"], xlab = "X5: Pressure (mmHg)", ylab = "Y: Appliances (Wh)",
     pch = 19, col = rgb(0,0,1,0.4), main = "X5 vs Y")
abline(lm(Y ~ X5, data = as.data.frame(src)), col = "red", lwd = 2)

par(op)


```
Variables X1 and X3 show positive linear relationships with Y. Variables X2,X4 and X5 show weak or no linear relationships with Y.

```{r}
# 5. Correlation matrix ----
cor_mat <- cor(df, use = "pairwise.complete.obs")
round(cor_mat, 3)
corrplot::corrplot(cor_mat, method = "ellipse", type = "upper", tl.cex = 0.8)
```

Correlations:

-   Strong positive correlation between X1 and X3 

-   Strong positive correlation between X2 and X4 

-   Moderate negative correlation between X2 and X5 

-   Moderate negative correlation between X4 and X5 

```{r}
# 6. Multicollinearity checks ----
X <- as.matrix(df[, c("X1","X2","X3","X4","X5")])
eigs <- eigen(cor(X))
cond_number <- sqrt(max(eigs$values) / min(eigs$values))
cond_number
lm_full <- lm(Y ~ X1 + X2 + X3 + X4 + X5, data = as.data.frame(df_num))
vifs <- car::vif(lm_full)
vifs
```
The condition number is 6.542, indicating weak multicollinearity overall in the entire model..
VIF > 5 already indicates moderate multicollinearity with  X1,X2,X3,X4 individually.
This also translates to being strong correlations observed earlier then predictors are almost linearly dependent.
This will cause instability in coefficient estimates and large standard errors.
Also model may being sensitive to small changes in data.
Appropriate Transformations are applied next to reduce multicollinearity and improve model stability.
```{r}
# 7. Suggested transformations (empirical + Box-Cox for Y) ----
if (all(df$Y > 0)) {
  bc <- MASS::boxcox(lm_full, plotit = FALSE)
  lambda_est <- bc$x[which.max(bc$y)]
  lambda_est
} else {
  lambda_est <- NA
  lambda_est
}
skewness_num <- function(x) {
  m3 <- mean((x - mean(x))^3)
  s3 <- sd(x)^3
  m3 / s3
}
skews <- sapply(df, skewness_num)
round(skews, 3)
```

```{r}
df_trans <- df
transform_notes <- list()
for (v in names(df)) {
  s <- skewness_num(df[[v]])
  if (s > 1) {
    if (all(df[[v]] > 0)) {
      df_trans[[v]] <- log(df[[v]])
      transform_notes[[v]] <- "log"
    } else {
      df_trans[[v]] <- sign(df[[v]]) * log(abs(df[[v]]) + 1)
      transform_notes[[v]] <- "signed_log"
    }
  } else if (s > 0.5) {
    if (all(df[[v]] >= 0)) {
      df_trans[[v]] <- sqrt(df[[v]])
      transform_notes[[v]] <- "sqrt"
    } else {
      df_trans[[v]] <- sign(df[[v]]) * sqrt(abs(df[[v]]))
      transform_notes[[v]] <- "signed_sqrt"
    }
  } else {
    transform_notes[[v]] <- "none"
  }
}
transform_notes
```

```{r}
# 8. Re-check correlations & VIF on transformed predictors ----
X_t <- as.matrix(df_trans[, c("X1","X2","X3","X4","X5")])
cor_mat_t <- cor(cbind(X_t, Y = df_trans$Y), use = "pairwise.complete.obs")
round(cor_mat_t, 3)
corrplot::corrplot(cor_mat_t, method = "ellipse", type = "upper", tl.cex = 0.8)
lm_full_t <- lm(Y ~ X1 + X2 + X3 + X4 + X5, data = df_trans)
vifs_t <- car::vif(lm_full_t)
vifs_t
```

